Dual host-parasite transcriptomes of apicomplexan Eimeria falciformis and its natural mouse host

### 2016-09-08 manuscript/bmc_article.tex
commit manuscript before starting to mess with including bibliography

%%% Load packages
\usepackage{amsthm,amsmath}
%\usepackage{asmath}
\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
%\usepackage{siunitx}
\usepackage[colorlinks]{hyperref} % To color blocks of text 
%\usepackage[dvipsnames]{xcolor} % More color options (see Wikibooks)  
\usepackage{changepage}
\usepackage{booktabs} % For prettier tables
\usepackage{float} % To control positions of e.g. tables or figures
%\usepackage[section]{placeins}
\usepackage{graphicx}
\usepackage[flushleft]{threeparttable}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	OVERVIEW READS PER SAMPLE, LAYOT AS EXPERIMENTAL OVERVIEW
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	OVERVIEW READS PER SAMPLE, LAYOT AS EXPERIMENTAL OVERVIEW
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We performed mRNA sequencing of caecum tissue from mice infected with
the apicomplexan parasite \textit{E. falciformis}. Oocysts and
sporozoites were included as ``environmental'' stages and processed
``\textit{in vitro}''. To follow the life cycle of the parasite, we
compare different time-points after infection.

We additionally used different mouse strains which display different
immunocompetence in infection trials measured by oocyst output to assess the influence of host
immunocompetence on parasite development (figure 1a,b). In these
experiments we analysed mice lacking the Recombination activating gene 1, Rag1, and compared them to the
parental C57BL/6 mouse line. Rag1 knock-out mice lack adaptive immune
response due to the absence of mature B and T lymphocytes.  All three
mouse strains were used for infections of both naive mice and previously
infected animals (onward referred to as "challenged").

Basic phenotyping showed differences in oocyst numbers
between the immunocompetent (C57BL/6) and
immunodeficient (Rag1-/-) host strain in infection of naive mice and 
in challenge infections. Immunocompetent NMRI mice were infected with a higher dosis  
of sporulated oocysts, and a drastic reduction of oocysts in faeces was seen in challenged mice
compared to naive mice (Figure 1a). Oocyst numbers in faeces peaked on days 8-9 and 
all mice had cleared the infection by day 14. Development of \textit{E. falciformis} intestinal stages is reflected by 
the percentage of parasite reads sequenced per time-point post infection (Figure 1b) and this was confirmed by quantitative reverse transcription PCR (RT qPCR) of parasitic 18S  (Figure 1c).

We thus use dual RNA-seq to analyse the life cycle of \textit{E. falciformis} under
the influence of host immune responses at early stages of infection. We used an experimental
design which allows to compare infections at 5 days post infection (dpi) for all experimental
conditions (NMRI, C57BL/6 and Rag1-/- mouse strains in naive and challenge infection). In addition,
additional time-points 3 dpi and 7 dpi were analysed from NMRI mice (Figure 1d).

\section*{Parasite and host dual transcriptomes can be assessed in parallel}
In this RNA-seq experiment, each replicate sample was enriched for caecum epithelial tissue which was pooled 
from three mice. mRNA was extracted and
sequencing libraries prepared. Two biological replicates were used for
all but two conditions (with one and three replicates, respectively). These libraries were sequenced on
several lanes of Illumina GAIIX (13 samples) and HiSeq machines (14 samples) and mapped to both
mouse and parasite genomes simultaneously to avoid spurious assignments
of reads in ultra conserved genomic regions. 
As samples and individual replicates were sequenced in batches to different depth and using different
instrumentation (Table 1) we performed quality controls (additional files xyz). These confirm
the absence of batch effects influencing analysis and quality of 
results.

Total numbers of sequenced reads as well as reads
mapped to either the \textit{E. falciformis} genome or the
mouse genome are indicated in Table 1 for all replicates. The number of total read mappings 
for individual replicates ranged from 11,479,260 (sample NMRI\_sporozoites\_rep2) to 145,033,326
(NMRI\_2ndInf\_5dpi\_rep1). At the latest time-point samples, 7 dpi, the overall mRNA
output of the sampled caecum tissue is dominated by parasite material with proportional mRNA
abundance of 77\% (in NMRI\_1stInf\_7dpi\_rep2) and 92\% (in
NMRI\_1stInf\_7dpi\_rep1). (Figure 1b).

\section*{Exclusion of samples with uncertain infection status}
A maximum of 86\% (sample NMRI\_1stInf\_7dpi\_rep1) and a minimum of
0.0326\% (sample NMRI\_2ndInf\_3dpi\_rep2) of mapped reads could be
assigned to the \textit{E. falciformis} genome in samples considered
infected (Table 1). We excluded samples NMRI\_2nd\_3dpi\_rep1 and NMRI\_2nd\_5dpi\_rep2
due to low parasite contribution to the overall transcriptome. 
Technically, this exclusion made it possible to obtain read counts in agreement with a 
negative binomial distribution (see additional file x). It is also likely that the number of reads in the excluded samples would have been
insufficient to fully normalise these datasets to those with the highest parasite contributions.
From a biological point of view, both excluded are samples from challenge
infection and we assume that the infection was cleared or reduced to a
non-detectable level. One sample (NMRI\_1stInf\_0dpi\_rep1) was excluded because
the uninfected control showed unexpected mapping of reads to the
\textit{E. falciformis} genome. We consider the three excluded samples to display an uncertain 
state of infection.

\section*{The mouse transcriptome changes upon \textit{E. falciformis} infection}
Statistical testing for differential expression between infected and
uninfected mice revealed changes in mRNA abundance becoming more
pronounced (involving more genes) at later time-points post infection (especially
at 7 dpi; Table 2). For mouse, 325 mRNAs were considered differently abundant (DA; FDR\<1\%)
between controls and 3 dpi, 1,804 mRNAs between controls and 5 dpi and 2,711
mRNAs between controls and 7 dpi. This lead to a combined set of 3,453 genes
responding to infection (Figure 2bi). DA mRNAs early in infection (3 dpi
and 5 dpi) were not a pure subset of genes DA later in
infection (7 dpi). Instead, the transcriptional profile of the mouse 
changes throughout the infection. Changes between controls and 7 dpi were in agreement with
previously published microarry data. Fold-change data obtained from
\textit{E. falciformis} infected mice at 6 dpi on Agilent microarrays 
(Schmidt et al 2012) or 7 dpi (our RNAseq data) against uninfected controls
show a strong correlation (Spearman's \rho = 0.74; Figure 2a). Considering both biological 
(e.g. exact time-points) and technical differences between the two methods this comparison 
confirms the adequacy of using dual RNA-seq for assessing the host transcriptome.

\section*{Epithelial responses depend on mouse immune status}
To further validate the pattern of a distinct
response in early infection we performed hierarchical clustering on
the (union of) mouse genes DA between different time-points post infection (Figure 2c).
By hierarchical clustering of mouse samples, three main sample clusters formed, “main clusters” 
(dendrogram of columns at top of Figure 2c). Immune deficient Rag1-/- mice cluster with control samples. 
There is no clear distinction between infected and non-infected Rag1-/- samples, which confirms the immune deficiency phenotype in these mice.

We identify a group of genes (cluster 4, Figure 2c) which change the transcriptional profile upon infection (all time-points) in immune competent mice only. In immune compromised Rag1-/- another group of genes (cluster 3, figure 2c) display the "infection profile" of immune competent mice also in control samples, indicating that these are mRNAs kept at a low abundance in a T and B cell dependent manner. Hence, these genes in some way depend on functional T and B cells also in controls (cluster 3) at the earliest time-point post infection (3 dpi, cluster 4). Gene Ontology, GO, terms enriched and regulated in cluster 3 are, e.g., "lipid modification" and "lipid metabolic process" and "protein intraciliary transport". Other enriched categories in cluster 3 are terms for regulation of different metabolic processes, for blood coagulation as well as terms containing "spinal cord", "axon" or "neuronal" regulation or development. (SI file x). We suggest that these are processes which are all intrinsically different in Rag1-/- mice compared to immune competent mice. The genes in cluster 4, which change only in immune competent mice, are highly abundant in controls including Rag1-/-. Infected immune competent mice have a lower abundance of these mRNAs, but in Rag1-/- there is no difference compared to controls. Therefore, these genes appear to be down-regulated upon infection by processes which depend on (mature) T and/or B cell activity, surprisingly also at 3 dpi. Enriched GO terms in this gene cluster include 13 terms for cytokines, particularly interleukin-1, IL-1, ("negative regulation of" and "secretion of"). Terms such as "negative regulation of viral (or inflammatory) response", "negative chemotaxis", "autophagy", "blood coagulation", "inositol phosphate-mediated signaling", and "positive regulation of calcineurin-NFAT" are also enriched in cluster 4. Inositol signaling can lead to release of calcium and calcineurin-dependent translocation of NFAT to the nucleus and activation of its target genes in T cells, but also many other cell types (reviewed by macian05).
GO enrichment of cluster 4 also highlights particular regulation of transforming growth factor \beta, TGF\beta, epidermal growth factor, EGF, and tumor necrosis factor, TNF. TGF\beta is important for wound healing in intestinal epithelium (beck03), and EGF regulates proliferation of epithelial cells and inhibits apoptosis (suzuki10). TNF appears to be dose-dependent and capable to suppress inflammatory responses (noti10) but is also reported to regulate proliferation of epithelial cells (kaiser97). Additionally, IL-1 and IL-6 are among the enriched GO terms in cluster 4. The IL-1 receptor (type I) is similar to Toll-like receptors and IL-1 induces innate immune responses in many cell types, and influence lymphocyte activity (dinarello09). IL-6 has been shown to support repair and inhibit apoptosis after epithelial wounding (kuhn14) and it is known to be important for development of Th17 responses (ref in kuhn intro). \\\\\\\\Th17 .... Stange's work (discuss Richard). IL-6 induces the Janus kinase, JAK, and signal transducer and activator of transcription STAT3, which, among other functions, is involved in wound healing in intestinal epithelial cells (pickert09). This analysis suggests that TGF\beta, TNF, EGF, IL-1 and IL-6 are main actors in the epithelial response to \textit{E. falciformis} infection and that the response is T and B cell dependent, also at early time-points. The pronounced changes late in infection (7 dpi) also reflect the expected onset of an adaptive immune response. This is underlined by enriched GO terms in gene clusters 1 and 5. Terms such as "antigen binding" and "immunoglobulin receptor binding" (molecular function, MF), and "immune system process", "adaptive immune response"and "innate immune response" (biological process, BP) are highly enriched in cluster 5 and confirm an activated adaptive immune response at this time-point. Among the same genes, natural killer cell regulation, JAK-STAT signalling, and IL-1 and interleukin-2, IL-2 production are enriched biological processes. IL-2 is one target of NFAT signaling and as mentioned above, JAK-STAT signaling can be induced by IL-6. This indicates that the early responses in, e.g., NFAT and IL-6 regulation induce distinct mRNA abundace differences later in infection (7 dpi).

Taken together, clusters 3 and 4 seem to capture the difference between Rag1-/- mice and immune competent mice in this parasitic infection. Some genes are different also between control animals (Rag1-/- and NMRI, cluster 3). These do not change upon infection in mice lacking mature T and B cells. Other genes (cluster 4) for which the transcriptional profile is similar in all controls, mRNA abundance also only changes upon infection only in animals with functional maturation of T and B cells. Using GO enrichment analysis, we show that regulation of IL-1, IL-6, TNF,TGF\beta and EGF are regulated in both first and challenge infections and suggest wound healing as a common function of this regulation, especially at 7 dpi.

\section{RNA processing is enriched in challenge infected immune competent mice} 
Three challenge infected samples (3 dpi, 5 dpi and 7 dpi) from immune competent mice show a distinct profile in cluster 6. This cluster is highly enriched for RNA processing and splicing, as well as terms for histone and chromatin modification. 
% daunorubicin metabolic process  --> cancer drug which prevents DNA replication by stabilizing DNA
% doxorubicin metabolic process --> cancer drug

@@

\section*{mRNA abundance .........}
We performed statistical tests to evaluate significant differences (FDR\<1\%) in mRNA abundance between different parasite life cycle stages, approximated by time post infection (Table 2). Between early time-points, 3 dpi and 5 dpi, 103 mRNAs were different, whereas between 3 dpi and 7 dpi 1399 mRNAs were DA, and between 5 dpi and 7 dpi 2084 mRNAs were DA (Figure 3a). This indicates that the major changes take place between 5 dpi and 7 dpi, and that variation is smaller between 3 dpi and 5 dpi. This motivated us to define 3 dpi and 5 dpi as "early infection" and 7 dpi as "late infection". Early and late infection samples were tested for DA compared to sporozoites and sporulated oocysts, resulting in 1697 and 3919 DA mRNAs, respectively.
%% Check venn vs table2 and change in ms + code
To evaluate this outcome further, we performed hierarchical clustering of (the union of) the DA mRNAs in the comparisons described above.

\section*{Major transcriptional differences in the \textit{E. falciformis} life cycle are independent of mouse immune status}
In the parasite transcriptome, we see no difference between infection in immune competent mice, or T and B cell deficient Rag1-/- mice, or between naive and challenge infected mice. This is surprising considering the measured differences in oocyst output in the same comparisons (Figure 1a), and the fact that these differences are visible in the mouse transcriptomes. On the parasite side, major patters instead seem to be determined by life cycle stages, independent of the host immune status. Distinct clusters of genes define early infection (3 dpi and 5 dpi) in which schizogony takes place, and separately, late infection (7 dpi) in which it is assumed that gametocytes are present. Extracellular (and in our experiments, in vitro) samples of sporozoites and oocysts also cluster separately and are defined by distinct gene clusters.

\section*{Early infection transcriptomes suggest parasite expansion}
%%%% Include cluster 3!!!!!!!
GO enrichment of the major sporozoite defining cluster with high mRNA abundance in this stage (cluster 4) proposes that different biosynthesis processes are important, as well as "maintenance of protein location in cell". Possibly, the latter is due to control of microneme protein localization as sporozoites prepare for invasion. Biosynthesis terms probably reflect preparation for asexual expansion and perhaps ATP production for motility in sporozoites, although no GO terms for motility are enriched in this cluster. In the following life cycle stages, measured 3 dpi and 5 dpi, in which several rounds of schizogony (asexual replication) take place, mRNAs in cluster 6 are abundant in all early samples (except one, discussed below). Among these "early infection" genes, several GO terms for biosynthetic acitivity are enriched, e.g., "ribosome biogenesis" and "cellular biosynthetic process", as well as terms for "gene expression" and RNA processes, including tRNAs and ncRNAs. "Cellular amino acid catabolic process" is also enriched. mRNAs in cluster 6 are low in abundance in both oocysts and late infection, whereas the sporozoite profile is less defined in this cluster. Therefore, this cluster is distinct for early infection with some features shared with invasive sporozoites. The enriched terms indicate replication and growth, highlighting the parasite's expansion in numbers on 3 dpi and 5 dpi, as supported both by previous knowledge about the life cycle and our increase in sequences from the parasite (Figure 1b).

\section*{Gametocyte presence likely determines transcriptome late in infection}
Two gene clusters have a distinct profile with high mRNA abundance on 7 dpi (clusters 2 and 7). Both clusters display low mRNA abundance in other life cycle stages, especially in oocysts and sporozoites. Enriched GO terms such as "movement of cell or subcellular component" and "microtubule-based movement" along with terms suggesting ATP production ("ATP generation from ADP") indicate the presence of motile and energy demanding gametocytes in these samples. Other cluster 2 terms such as "chitin metabolic process" along with enriched metabolic and biosynthetic processes and "gene expression" in cluster 7 suggests start of or preparation for encystation at 7 dpi. In addition, cluster 2 is enriched for a number of GO terms for "blood coagulation" and reflect 
%%%(double check) 
the presence of Thrombospondin type I domains in the protein products of cluster 2 mRNAs. Thrombospondin type 1 domains have been reported in \textit{E. tenella} microneme
localizing proteins, MIC, e.g. MIC4 (Tomley01). MIC4 mRNA was reported in sporozoites where it localizes to the apical
end, and in late schizonts and late oocyst stages, when sporozoites are forming. This suggests that \textit{E. falciformis} prepares for invasion already in gametocytes or during oocyst formation, however this is speculative.

\section*{Oocysts...}
Oocysts have strong high or low mRNA abundance profiles and few intermediate abundance mRNAS. Clusters 1 and 5 contain mRNAS with high abundance in oocysts and all other clusters have low abundance in this stage. GO enrichment in cluster 1 contains only one term (adj. p-value 0.11) for "DNA-templated transcription, initiation". Cluster 5 is enriched for terms related to stress responses, "DNA repair", "protein modification process" and for "cell differentiation". Stress responses and DNA repair can be a result of storage in potassium dichromate of mouse faeces with oocysts before purification. Initiation of transcription and cell differentiation could reflect slow preparation for invasion when oocysts are taken out of storage and purified for RNA extraction. Overall, the oocyst profile in five of 7 gene clusters is characterized by below average abundance of mRNAs, as can be expected in this life cycle stage suited for long-term survival outside of the host. Taken together, sample and gene clustering indicates that genes in clusters 6 are abundant only early in infection and are candidates for merozoite specific genes, clusters 2 and 7 might be useful to characterise gametocytes. Genes in cluster 4 can be evaluated as sporozoite specific genes, and clusters 1 and 5 as oocyst specific genes in \textit{E. falciformis}.

>>>>

%%% Load packages
\usepackage{amsthm,amsmath}
\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
%\usepackage{siunitx}
\usepackage[colorlinks]{hyperref} % To color blocks of text 
%\usepackage[dvipsnames]{xcolor} % More color options (see Wikibooks)  
\usepackage{changepage}
\usepackage{booktabs} % For prettier tables
\usepackage{float} % To control positions of e.g. tables or figures
%\usepackage[section]{placeins}
\usepackage{graphicx}
\usepackage[flushleft]{threeparttable}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	OVERVIEW READS PER SAMPLE, LAYOT AS EXPERIMENTAL OVERVIEW
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We performed mRNA sequencing of caecum epithelial tissue from mice infected with
the apicomplexan parasite \textit{E. falciformis}. Oocysts and
sporozoites were included as "environmental" stages and processed
in vitro. To follow the life cycle of the parasite, we
compare different time-points after infection. We additionally used different mouse strains which display different immunocompetence in infection trials, measured by oocyst output in order to assess the influence of host immunocompetence on parasite development (Figure 1a and 1b). Immune competent NMRI mice were infected and sampled at three time-points post infection, p.i. We also infected mice lacking the Recombination activating gene 1, Rag1, (Rag1-/-) and compared them to the parental C57BL/6 mouse line. Rag1-/- mice lack mature B and T lymphocytes, which is taken as a proxy for absent adaptive immunity. All three mouse strains were used for infections of both naive mice and previously infected (and recovered) animals (onward referred to as "challenged" or "challenge infection").
\newline Basic phenotyping (Figure 1a) showed differences in oocyst numbers between the immunocompetent (C57BL/6) and immunodeficient (Rag1-/-) host strains in naive and challenge infected mice. Immunocompetent NMRI mice were infected with a higher dosis of sporulated oocysts, and a drastic reduction of oocysts in faeces was seen in challenged mice compared to naive mice. Oocyst numbers in faeces peaked on days 8-9 andall mice had cleared the infection by day 14. Development and replication of \textit{E. falciformis} intestinal stages is reflected by the percentage of parasite reads sequenced per time-point post infection (Figure 1b) and this was confirmed by quantitative reverse transcription PCR (RT qPCR) of parasitic 18S (Figure 1c).
\newline We thus use dual RNA-seq to analyse the life cycle of \textit{E. falciformis} under
the influence of host immune responses at early and late stages of infection. We used an experimental
design which allows to compare infections at 5 days post infection (dpi) for all experimental
conditions (NMRI, C57BL/6 and Rag1-/- mouse strains in naive and challenge infection). Additional time-points 3 dpi and 7 dpi were analysed from NMRI mice (Figure 1d).

\section*{Parasite and host dual transcriptomes can be assessed in parallel}
In this RNA-seq experiment, each replicate sample was enriched for caecum epithelial tissue and pooled 
from three mice. mRNA was extracted and sequencing libraries prepared. Two biological replicates were used for all but two conditions (with one and three replicates, respectively). Libraries were sequenced on several lanes of Illumina GAIIX (13 samples) and HiSeq machines (14 samples) and mapped to both mouse and parasite genomes simultaneously to avoid spurious assignments of reads in ultra conserved genomic regions. As samples and individual replicates were sequenced in batches to different depth and using different instrumentation (Table 1) we performed quality controls (additional files xyz). These confirm the absence of batch effects influencing analysis and quality of results.

Total numbers of sequenced reads as well as reads mapped to either the \textit{E. falciformis} genome or the mouse genome are indicated in Table 1 for all replicates. The number of total read mappings for individual replicates ranged from 25,362,739 (sample Rag\_1stInf\_0dpi\_rep1) to 139,749,046 (NMRI\_1stInf\_7dpi\_rep2). At the latest time-point samples, 7 dpi, the overall mRNA output of the sampled caecum tissue is dominated by parasite material with proportional mRNA abundance of 77\% (in NMRI\_1stInf\_7dpi\_rep2) and 92\% (in NMRI\_1stInf\_7dpi\_rep1). (Figure 1b).

\section*{Exclusion of samples with uncertain infection status}
A maximum of 92\% (sample NMRI\_1stInf\_7dpi\_rep1) and a minimum of 0.064\% (sample NMRI\_2ndInf\_3dpi\_rep2) of mapped reads could be assigned to the \textit{E. falciformis} genome in samples considered infected (Table 1). We excluded samples NMRI\_2nd\_3dpi\_rep1 and NMRI\_2nd\_5dpi\_rep2 due to low parasite contribution to the overall transcriptome. Technically, this exclusion made it possible to obtain read counts in agreement with a negative binomial distribution (see additional file x). It is also likely that the number of reads in the excluded samples would have been insufficient to fully normalise these datasets to those with the highest parasite contributions. From a biological point of view, both excluded samples are samples from challenge infection and we assume that the infection was cleared or reduced to a non-detectable level. One sample (NMRI\_1stInf\_0dpi\_rep1) was excluded because the uninfected control showed unexpected mapping of reads to the \textit{E. falciformis} genome. We consider the three excluded samples to display an uncertain state of infection.

\section*{The mouse transcriptome changes upon \emph{E. falciformis} infection}
Statistical testing for differential expression between infected and uninfected mice revealed changes in mRNA abundance becoming more pronounced (containing more genes) at later time-points post infection (Table 2). For mouse, 325 mRNAs were considered differently abundant (DA; FDR < 0.01) between controls and 3 dpi, 1,804 mRNAs between controls and 5 dpi and 2,711 mRNAs between controls and 7 dpi. This lead to a combined set of 3,453 unique genes responding to infection (Figure 2bi). DA mRNAs early in infection (3 dpi and 5 dpi) were not a strict subset of genes DA later in infection (7 dpi). Instead, the transcriptional profile of the mouse changes throughout the infection. Differences between controls and 7 dpi were in agreement with previously published microarry data. Fold-change data obtained from \textit{E. falciformis} infected mice at 6 dpi on Agilent microarrays (\cite{schmidt12}) or 7 dpi (our RNA-seq data) against uninfected controls show a strong correlation (Spearman's $\rho$ = 0.74; Figure 2a). Considering both biological differences in the experiments (e.g. exact time-points for samples), and technical differences between the two methods this comparison confirms the adequacy of using dual RNA-seq for assessing the host transcriptome.

\section*{Epithelial responses vary with mouse immune status}
A distinct response in early infection is indicated by the number of DA mRNAs between different time-points. To further validate the pattern we performed hierarchical clustering on the (union of) mouse genes DA between different time-points post infection (Figure 2c). By hierarchical clustering of mouse samples, three main sample clusters formed (dendrogram of columns at top of Figure 2c). Immune deficient Rag1-/- mice cluster with control samples. There is no clear distinction between infected and non-infected Rag1-/- samples, which confirms the immune deficiency phenotype in these mice. 
We identify a group of genes (cluster 4, Figure 2c) which changes transcriptional profile upon infection (all time-points of infected samples) in immune competent mice only. In immune compromised Rag1-/-, another group of genes (cluster 3, figure 2c) display the "infection profile" of immune competent mice also in control samples, indicating that these are mRNAs kept at a low abundance in healthy animals in a T and B cell dependent manner. Hence, these genes in some way depend on functional T and B cells also in controls (cluster 3) and at the earliest time-point post infection (3 dpi, cluster 4). 
Gene Ontology, GO, terms enriched in cluster 3 are, e.g., "lipid modification" and "lipid metabolic process" and "protein intraciliary transport". Other enriched terms in cluster 3 are terms for regulation of other metabolic processes, for blood coagulation as well as terms containing "spinal cord", "axon" or "neuronal" regulation or development. (SI file x). We suggest that these are processes which are all intrinsically different in Rag1-/- mice compared to immune competent mice. The genes in cluster 4, which change only in immune competent mice, are highly abundant in controls including Rag1-/-. Infected immune competent mice have a lower abundance of these mRNAs.
 Therefore, these genes appear to be down-regulated upon infection by processes which depend on (mature) T and/or B cell activity, surprisingly also at 3 dpi. Enriched GO terms in this gene cluster include 13 terms for cytokines, particularly interleukin-1, IL-1, ("negative regulation of" and "secretion of"). Terms such as "negative regulation of viral (or inflammatory) response", "negative chemotaxis", "autophagy", "blood coagulation", "inositol phosphate-mediated signaling", and "positive regulation of calcineurin-NFAT" are also enriched in cluster 4. Inositol signaling can lead to release of calcium and calcineurin-dependent translocation of NFAT to the nucleus and activation of its target genes in T cells, but also many other cell types (reviewed by macian05). GO enrichment of cluster 4 also highlights particular regulation of transforming growth factor$\beta$, TGF$\beta$, epidermal growth factor, EGF, and tumor necrosis factor, TNF. TGF$\beta$ is important for wound healing in intestinal epithelium (beck03), and EGF regulates proliferation of epithelial cells and inhibits apoptosis (suzuki10). TNF appears to be dose-dependent and capable to suppress inflammatory responses (noti10) but is also reported to regulate proliferation of epithelial cells (kaiser97). Additionally, IL-1 and IL-6 are among the enriched GO terms in cluster 4. The IL-1 receptor (type I) is similar to Toll-like receptors and IL-1 induces innate immune responses in many cell types, and influences lymphocyte activity (dinarello09). IL-6 has been shown to support repair and inhibit apoptosis after epithelial wounding (kuhn14) and it is known to be important for development of Th17 responses (ref in kuhn intro). 
%Th17 .... Stange's work (discuss Richard). 
IL-6 induces the Janus kinase, JAK, and signal transducer and activator of transcription STAT3, which, among other functions, is involved in wound healing in intestinal epithelial cells (pickert09). This analysis suggests that TGF$\beta$, TNF, EGF, IL-1 and IL-6 are main actors in the epithelial response to \textit{E. falciformis} infection and that the response is T and B cell dependent, also at early time-points. Enrichment of these biological processes further indicates that hosts invest resources in intestinal healing also at early time-points of \textit{E. falciformis} infection.

\newline The pronounced changes late in infection (7 dpi) reflect the expected onset of an adaptive immune response. This is underlined by enriched GO terms in gene clusters 1 and 5. Terms such as "antigen binding" and "immunoglobulin receptor binding" (molecular function, MF), and "immune system process", "adaptive immune response"and "innate immune response" (biological process, BP) are highly enriched in cluster 5 and confirm an activated adaptive immune response at this time-point. Among the same genes, natural killer cell regulation, JAK-STAT signalling, and IL-1 and interleukin-2, IL-2 production are enriched biological processes. IL-2 is one target of NFAT signaling and as mentioned above, JAK-STAT signaling can be induced by IL-6. This indicates that the early responses in, e.g., NFAT and IL-6 regulation induce distinct mRNA abundace differences later in infection (7 dpi).

Taken together, clusters 3 and 4 highlight the difference between Rag1-/- mice and immune competent mice in this parasitic infection. Some genes are different also between control animals (Rag1-/- and NMRI, cluster 3). These do not change upon infection in mice lacking mature T and B cells. Other genes (cluster 4) for which the transcriptional profile is similar in all controls, mRNA abundance also changes upon infection only in animals with functional maturation of T and B cells. Using GO enrichment analysis, we show that regulation of IL-1, IL-6, TNF,TGF$\beta$ and EGF are regulated in both first and challenge infections and suggest wound healing as a probably function of this regulation, especially at 7 dpi.

\section*{RNA processing is enriched in challenge infected immune competent mice} 
Three challenge infected samples (3 dpi, 5 dpi and 7 dpi) from immune competent mice show a distinct profile in cluster 6. This cluster is highly enriched for GO terms for RNA processing and splicing, as well as terms for histone and chromatin modification. \ldots
\newline
Include or not?
- daunorubicin metabolic process > cancer drug which prevents DNA replication by stabilizing DNA
- doxorubicin metabolic process  > cancer drug
\newline

@@

\section*{Transcriptional differences in the parasite life cycle are independent of mouse immune status}
We performed statistical tests to evaluate significant differences (FDR$<$0.01) in mRNA abundance between different parasite life cycle stages, approximated by time post infection (Table 2). Between early time-points, 3 dpi and 5 dpi, 103 mRNAs were different, whereas between 3 dpi and 7 dpi 1399 mRNAs were DA, and between 5 dpi and 7 dpi 2084 mRNAs were DA (Figure 3a). This indicates that the major changes take place between 5 dpi and 7 dpi, and that variation is smaller between 3 dpi and 5 dpi. This motivated us to define 3 dpi and 5 dpi as "early infection" and 7 dpi as "late infection". Early and late infection samples were tested for DA compared to sporozoites and sporulated oocysts, resulting in 1697 and 3919 DA mRNAs, respectively.
%% Check venn vs table2 and change in ms + code
To evaluate this outcome further, we performed hierarchical clustering of (the union of) the DA mRNAs in the comparisons described above and applied GO enrichment analysis of gene clusters (annotations are inferred from orthologs in other \textit{Eimeria} spp. or \textit{T. gondii}). In the parasite transcriptome, we see no difference between infection in immune competent mice, or T and B cell deficient Rag1-/- mice, or between naive and challenge infected mice. This is surprising considering the measured differences in oocyst output in the same comparisons (Figure 1a), and the fact that these differences are visible in the mouse transcriptomes. On the parasite side, major patters instead seem to be determined by life cycle stages, independent of the host immune status. Distinct clusters of genes define early infection (3 dpi and 5 dpi) in which schizogony (asexual reproduction) takes place, and separately, late infection (7 dpi) in which it is assumed that gametocytes are present. Extracellular samples of sporozoites and oocysts also cluster separately and are defined by distinct gene clusters.

\section*{Early infection transcriptomes reflect parasite expansion}
GO enrichment of the major sporozoite defining cluster with high mRNA abundance in this stage (cluster 4) proposes that different biosynthesis processes are important, as well as "maintenance of protein location in cell". Possibly, the latter is due to control of microneme protein localization as sporozoites prepare for invasion. Biosynthesis terms probably reflect preparation for asexual expansion and perhaps ATP production for motility in sporozoites, although no GO terms for motility are enriched in this cluster. In the following intracellular life cycle stages, measured at 3 dpi and 5 dpi, in which several rounds of schizogony (asexual replication) take place, mRNAs in cluster 6 are abundant in all early samples (except one, discussed below). Among these "early infection" genes in cluster 6, several GO terms (biological process) for biosynthetic acitivity are enriched, e.g., "ribosome biogenesis" and "cellular biosynthetic process", as well as terms for "gene expression" and RNA processing, including terms for tRNAs and ncRNAs. "Cellular amino acid catabolic process" is also enriched. In cluster 3, with a similar profile as cluster 6, no GO terms are enriched. Cluster 6 is distinct in early infection with a biosynthesis profile shared with invasive sporozoites. Replication and growth-related processes being enriched highlights the parasite's expansion in numbers on 3 dpi and 5 dpi, as supported both by previous knowledge about the life cycle and our increase in sequences from the parasite (Figure 1b).

\section*{Gametocytes likely determine transcriptome late in infection}
Two gene clusters have a distinct profile with high mRNA abundance on 7 dpi (clusters 2 and 7). Both clusters display low mRNA abundance in other life cycle stages, especially in oocysts and sporozoites. Enriched GO terms such as "movement of cell or subcellular component" and "microtubule-based movement" along with terms suggesting ATP production ("ATP generation from ADP") indicate the presence of motile and energy demanding gametocytes in these samples. Other cluster 2 terms such as "chitin metabolic process" along with enriched metabolic and biosynthetic processes and "gene expression" in cluster 7 suggests ongoing encystation at 7 dpi, fitting the timing of oocyst output which peaks at 8-9 dpi. In addition, cluster 2 is enriched for a number of GO terms for "blood coagulation" and reflect 
%%%(double check) 
the presence of Thrombospondin type I domains in the protein products of cluster 2 mRNAs. Thrombospondin type 1 domains have been reported in \textit{E. tenella} microneme localizing proteins, MIC, e.g. MIC4 (Tomley01). MIC4 mRNA was reported in sporozoites where it localizes to the apical end, and in late schizonts and late oocyst stages, when sporozoites are forming. This suggests that \textit{E. falciformis} prepares for invasion already in gametocytes or during oocyst formation, however this is speculative.

\section*{Oocysts are characterized by stress responses and differentiation}
Clusters 1 and 5 contain mRNAs with high abundance in oocysts and all other clusters have low abundance in this stage. GO enrichment in cluster 1 contains only one term (adj. p-value 0.11) for "DNA-templated transcription, initiation". Cluster 5 is enriched for terms related to stress responses, "DNA repair", "protein modification process" and for "cell differentiation". Stress responses and DNA repair can be a result of storage in potassium dichromate of mouse faeces with oocysts before purification. Initiation of transcription and cell differentiation probably reflects (slow) preparation for invasion when oocysts are taken out of storage and purified for RNA extraction. Overall, the oocyst profile in five out of seven gene clusters is characterized by below average abundance of mRNAs, as can be expected in this life cycle stage which endures long-term survival outside the host. Taken together, sample and gene clustering indicates that genes in clusters 6 are abundant only early in infection and are candidates for merozoite specific genes, whereas clusters 2 and 7 might be useful to characterise gametocytes. Genes in cluster 4 can be considered sporozoite specific, and clusters 1 and 5 as oocyst specific genes in \textit{E. falciformis}.
